---
title: "Week2InClass"
output: html_document
date: "2024-01-30"
---

```{r}
#read in and look at the data
load("nepal.anthro.rdata")
names(nepal.anthro)

#num variable tells the visit number. longitudinal study, 5 visits, just take the first visit
d= nepal.anthro %>% select(.,age,sex,ht,wt,arm,num) %>% filter(.,num==1)
d <- d[complete.cases(d),] # drop cases without one or more of these variables
d <- d[order(d$age),-6] # Sort the data by age and drop "num"
dim(d)
```
```{r}
par(mar=c(1,1,1,1)) #modification to the R plotting default setting. how many lines want in margin region. 
pairs(select(d,age,sex,ht,wt,arm),pch=".")
#Display key variables --> Pairwise Scatterplot. 
#Row is in Y axis. Variable in the column is the X variable.

#2,1 / 3,1  / 4,1
#ht vs age, wt vs age, 
```
```{r}
#Simple Scatterplot --> How arm circumference changes with age
ggplot(d, aes(x = age, y = arm)) +
geom_jitter(alpha = 0.5) +
theme_bw() +
scale_y_continuous(breaks=seq(8,18,2),limits=c(8,18)) +
scale_x_continuous(breaks=seq(0,60,6),limits=c(0,60)) +
labs(y = "Arm cirumference (in cm)", x = "Age (in months)") +
theme(panel.border = element_blank(),
panel.grid.major = element_blank(),
panel.grid.minor = element_blank(),
axis.line = element_line(colour = "black")) 
```

## Define, fit and interpret a linear spline model
Start with knots at 6, 12, 24, 36, 48 months and then drop unnecessary knot points. 
```{r}
d=mutate(d,
agesp6=ifelse(age-6>0, age-6,0), #if age-6>0, assign age-6, otherwise 0 
agesp12=ifelse(age-12>0, age-12,0),
agesp24=ifelse(age-24>0, age-24,0),
agesp36=ifelse(age-36>0,age-36,0),
agesp48=ifelse(age-48>0, age-48,0)
)
# check what predictors for linear splines look like

pairs(select(d,age,agesp6,agesp12,agesp24,agesp36, agesp48),pch=".",
main="Pairs Plot of Age Linear Spline Vars")

#first column --> basis function. for changing the slope at each knot. 
```
```{r}
cc=complete.cases(select(d,age,arm))
d.cc=filter(d,cc)
d.cc = arrange(d.cc,age)
reg1<-lm(data=d.cc, arm~age+agesp6+agesp12+agesp24+agesp36+agesp48)
#fit this linear spline. y ~ mean model specification, absent the intercept
summary.lm(reg1)  
```
y is the arm circumference
i = 1, 2 ..., 185 = n
Yi = B0+B1axi+B2(axi-6)^+ +B3(axi-12)^+ B4(axi-24)^+ B5(axi-36)^+ + B6(axi-48)^+ + ei
ei ~ iid N(0, sigma^2)
^^ This is the mathematical representation, be sure to write ei. 

```{r}
#residual standard error is actually residual standard deviation. estimate of the sigma. population parameter. 
```

*Value of the intercept:*   
on average, children at birth has an arm circumference of 10.94 cm. (give 95% confidence interval)
10.94 +- 1.96*0.55

*Coefficient of Age:*  
Among children under 6 months of age, the average arm circumference differs by 0.38cm when comparing children who differ in age by 1 month.

*Coefficient of Age6:*  
`B hat 1`+`B hat 2` 
the mean arm circumference for 12 months old --> B0 + B1* 12 + B2 * 6 --> 10.94 + 12 * 0.38 + 6 * -0.43. -> mean estimate. 
standard deviation -> just the residual standard deviation. 

Because there is little reason to think that mean AC goes down between ages 6 and 12 months and because
there is little evidence in the data supporting changes in slope after 12 months, let’s fit a second model with
only two slope changes at 6 and 12 months, then a third with only a break at 6 months

```{r}
reg2<-lm(data=d.cc, arm~age+agesp6+agesp12)
reg3<-lm(data=d.cc, arm~age+agesp6)
summary.lm(reg2); summary.lm(reg3)
```

Make a figure of the observed data and the 3 linear spline models.
```{r} 
ggplot(d.cc, aes(x = age, y = arm)) + theme_bw() +
geom_jitter(alpha = 0.5) +
geom_line(aes(x = age, y = reg1$fitted.values),color="black") +
geom_line(aes(x = age, y = reg2$fitted.values),color="red") +
geom_line(aes(x = age, y = reg3$fitted.values),color="blue") +
scale_y_continuous(breaks=seq(8,18,2),limits=c(8,18)) +
scale_x_continuous(breaks=seq(0,60,6),limits=c(0,60)) +
labs(y = "Arm cirumference (in cm)", x = "Age (in months)") +
theme(panel.border = element_blank(),
panel.grid.major = element_blank(),
panel.grid.minor = element_blank(),
axis.line = element_line(colour = "black"))
```


# Lecture 4 Handout
```{r}
d= nepal.anthro %>% select(., arm,age,sex,ht,num) %>% filter(.,num==1)
d <- d[complete.cases(d),] # drop cases without one or more of these variables
d <- d[order(d$age),-6]
```


## Does the AC vs. age relationship vary by effect modifier sex?
1. Fit the linear spline model, separately for each sex
```{r}
d$agesp6 = ifelse(d$age-6>0, d$age-6,0)
reg3.male = lm(arm~age+agesp6, data=d, subset=sex==1)
reg3.female = lm(arm~age+agesp6, data=d, subset=sex==2)
cbind(summary(reg3.male)$coefficients[,1:2],
      summary(reg3.female)$coefficients[,1:2])
#male on the left, female on the right. 
```
To fit the model, create interaction terms in the model to allow for a separate AC and age association for each sex. 
```{r}

d= nepal.anthro %>% select(., arm,age,sex,ht,num) %>% filter(.,num==1)
d <- d[complete.cases(d),] # drop cases without one or more of these variables
d <- d[order(d$age),-6] # Sort the data by age and drop "num"

d$agesp6=ifelse(d$age-6>0, d$age-6,0)
reg3.male=lm(arm~age+agesp6,data=d,subset=sex==1)
reg3.female=lm(arm~age+agesp6,data=d,subset=sex==2)
cbind(summary.lm(reg3.male)$coefficients[,1:2],
summary.lm(reg3.female)$coefficients[,1:2])

d$int.female.age=d$sex*d$age
d$int.female.agesp6=d$sex*d$agesp6
reg4=lm(data=d,arm~female + age + agesp6 + int.female.age + int.female.agesp6)
summary.lm(reg4)$coeff

```



