---
title: "140.653 Lab 03: Adjusted Variable Plots"
subtitle: | 
  | Methods in Biostatistics 3 (140.653) 
  | Jason Haw, Elizabeth Colantuoni and Shuai Li
author: "Jason Haw, Elizabeth Colantuoni and Shuai Li"
output: 
  rmdformats::robobook:
      number_sections: yes
      highlight: haddock
header-includes:
- \usepackage[T1]{fontenc}
- \usepackage{fancyhdr}
- \usepackage{amsmath}
- \usepackage{float}
- \floatplacement{figure}{H}
mainfont: Arial
---

<style>

body {
    font-family: Arial, sans-serif;
}

.book .book-body .page-inner section.normal h1 {
  font-size: 24px;
  font-family: Arial, sans-serif;
}

.book .book-body .page-inner section.normal h1.title {
  font-size: 2em;
  margin-top: 0;
  color: #024873;
}

.book .book-body .page-inner section.normal h1.subtitle {
    font-size: 1em;
    font-weight: 400;
    margin-top: -15px;
    color: #024873;  
}


.book .book-body .page-inner section.normal h2, 
.book .book-body .page-inner section.normal h3, 
.book .book-body .page-inner section.normal h4 {
  font-size: 20px;
  font-family: Arial, sans-serif;  
}

</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=5)       # To override default scientific notation for large numbers
options(digits=3)
```

At the end of this lab, you should be able to:

1. Describe how an adjusted variable regression is similar to multiple linear regression
2. Create an adjusted variable plot from a multiple linear regression

<br>

We will use the Nepali anthropometry data set provided in this course. We will use the following variables:

* `arm`: arm circumference in cm
* `age`: age in months
* `ht`: height in cm
* `wt`: weight in kg
* `sex`: sex at birth (male = 1, female = 2)

**In this lab, our goal is to visualize the relationship between arm circumference ($Y_i$) and age ($X_i$) after adjusting for height, weight, and sex ($\mathbf{Z}_i$).**

# Adjusted Variable Plots, Explained

An adjusted variable plot is a graphical tool that allows us to visualize a specific coefficient for a variable of interest in a multiple linear regression model. We "adjust" the outcome variable $Y_i$ and variable of interest $X_i$ for the other variables ($\mathbf{Z}_i$) and visualize that adjustment as a scatterplot.

The adjustment is done in such a way that we remove the information from the other variables (e.g. $\mathbf{Z}_i$) from both $Y_i$ and $X_i$ and examine what remains. This is done by regressing the residuals of a linear model of $Y_i$ regressed on $\mathbf{Z}_i$ on the residuals of a linear model of $X_i$ regressed on $\mathbf{Z}_i$.

This implies that if a simple linear regression was run on the scatterplot of these residuals (residuals for $Y_i$ verses residuals for $X_i$), the linear slope in that plot corresponds to the coefficient of $X_i$ in the multiple linear regression model.

To understand how this works mathematically, consider a multiple linear regression model with a single covariate defining $Z_i$:

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + e_i$$

A sketch of the proof follows:

We can express the relationship of $Y_i$ and $X_i$ only in terms of $Z_i$ and derive their corresponding residuals:

$$Y_i = \alpha_0 + \alpha_1 Z_i + e_{iZY} \implies e_{iZY} = Y_i - (\alpha_0 + \alpha_1 Z_i)$$

$$X_i = \gamma_0 + \gamma_1 Z_i + e_{iZX} \implies e_{iZX} = X_i - (\gamma_0 + \gamma_1 Z_i)$$
The "adjustment" we make on $Y_i$ and $X_i$ is taking their corresponding residuals $e_{iZY}$ and $e_{iZX}$, i.e. the residuals represent the information about $Y_i$ and $X_i$ that is not explained by $Z_i$. We can then regress the residuals as a simple linear regression:

$$e_{iZY} = \beta_1 e_{iZX} + e_i$$
$$Y_i - (\alpha_0 + \alpha_1 Z_i) = \beta_1 [X_i - (\gamma_0 + \gamma_1 Z_i)] + e_i $$
Rearranging the terms:

$$Y_i = \alpha_0 + \beta_1 \gamma_0 + \beta_1 X_i + (\alpha_1 + \beta_1 \gamma_1) Z_i + e_i $$
We see that:

* $\beta_0 = \alpha_0 + \beta_1 \gamma_0$
* $\beta_1 = \beta_1$
* $\beta_2 = \alpha_1 + \beta_1 \gamma_1$

The adjusted variable regression and multiple linear regression give the same coefficient $\beta_1$ for $X_i$.


# Creating an Adjusted Variable Plot

## One Additional Covariate

Our goal in this section is to visualize the relationship between arm circumference ($Y$) and age ($X$) after adjusting for height ($Z$)

We first prepare our data for analysis

```{r dataprep, warning = FALSE, message = FALSE}
library(tidyverse)

# A note about loading data in R
# When loading data in either .rds or .rdata format, first locate the
# file on your system and copy the complete file path.
# Here, we loaded the data set on the same directory where the R project
# was created, so we only need the file name itself
# Since you are downloading this Rmd file without the R project where this was
# originally housed, you need to change the load() function below
# to the FULL path of the file on your system
# Note that for Windows, you need to use backslashes \ to identify subfolders
# as the default frontslash / does not work in R

load("nepal.anthro.rdata")

data1 <- nepal.anthro |> 
  select(c(1:16)) |> 
  filter(num == 1) |> 
  filter(!is.na(age) & !is.na(arm)) |>
  arrange(age)
```

The steps for the regression are as follows:

1. Obtain residuals $R(Y|Z)$ from the regression model of arm circumference ($Y$) on height ($Z$).
2. Obtain residuals $R(X|Z)$ from the regression model of age ($X$) and height ($Z$).
3. Plot $R(Y|Z)$ and $R(X|Z)$. The slope coefficient is equivalent to the coefficient of $X$ in a multiple regression model of $Y$ on $X$ and $Z$.


```{r onecov, warning = FALSE, message = FALSE}
# Obtain residuals R(Y|Z) and R(X|Z)
data1 <- data1 |>
  mutate(resid.arm1 = lm(arm ~ ht, data = data1)$residuals,
         resid.age1 = lm(age ~ ht, data = data1)$residuals)

# Run the model of R(Y|Z) on R(X|Z)
resid.model1 <- lm(resid.arm1 ~ resid.age1, data = data1)
summary(resid.model1)

# We see that the slope in resid.model1 is the same 
# as the beta of age in the multiple regression model
mr1 <- lm(arm ~ age + ht, data = data1)
summary(mr1)

# Create custom theme
custom_theme <- theme(
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  axis.text = element_text(size = 12),
  axis.title = element_text(size = 12, face = "bold"),
  axis.line = element_line(size = 0.5),
  plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
)

# Plot the estimates and we add a fitted line
avplot1 <- ggplot(data = data1,
                  aes(x = resid.age1, y = resid.arm1)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(aes(y = predict(resid.model1)), method = "lm", formula = y ~ x,
              linewidth = 1, color = "#024873", se = FALSE) +
  labs(x = "Residuals of age on height",
       y = "Residuals of arm circumference on height",
       title = "Adjusted variable plot for arm circumference on age adjusting for height") +
  custom_theme
avplot1
```

## Two Covariates (or More)

We implement a similar procedure for more than one covariate. Here, we visualize the relationship of arm circumference ($Y$) and age ($X$), adjusting for height, weight, and sex ($\mathbf{Z}$).

```{r twocov, warning = FALSE, message = FALSE}
# Obtain residuals R(Y|Z) and R(X|Z)
data1 <- data1 |>
  mutate(resid.arm2 = lm(arm ~ ht + wt + sex, data = data1)$residuals,
         resid.age2 = lm(age ~ ht + wt + sex, data = data1)$residuals)

# Run the model of R(Y|Z) on R(X|Z)
resid.model2 <- lm(resid.arm2 ~ resid.age2, data = data1)
summary(resid.model2)

# We see that the slope in resid.model1 is the same 
# as the beta of age in the multiple regression model
mr2 <- lm(arm ~ age + ht + wt + sex, data = data1)
summary(mr2)

# Plot the estimates and we add a fitted line
avplot2 <- ggplot(data = data1,
                  aes(x = resid.age2, y = resid.arm2)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(aes(y = predict(resid.model2)), method = "lm", formula = y ~ x,
              linewidth = 1, color = "#024873", se = FALSE) +
  labs(x = "Residuals of age on height, weight, age",
       y = "Residuals of arm circumference on height, weight, age",
       title = "Adjusted variable plot for arm circumference on age \n adjusting for height, weight, age") +
  custom_theme
avplot2
```


# In-Class Exercise

With the same data above, using the observed age ($X$), sex ($Z_1$) height($Z_2$), simulate the arm circumference value ($Y$) from the following regression models:

a. $Y_i = 7.8 + 0 \cdot X_i + 0.03Z_{1i} + 0.07Z_{2i} + \epsilon_i, \epsilon \sim N(0, 0.93)$
b. $Y_i = 7.8 + 0.035 \cdot X_i + 0.03Z_{1i} + 0.07Z_{2i} + \epsilon_i, \epsilon \sim N(0, 0.93)$
c. $Y_i = 7.8 + 0.05 \cdot X_i - 0.025(X_i - 6)I(X_i > 6) + 0.03Z_{1i} + 0.07Z_{2i} + \epsilon_i, \epsilon \sim N(0, 0.93)$ (Note that $I(\cdot)$ is an indicator function).

For each of these simulated data sets, create an adjusted variable plot of arm circumference on age, adjusting for height and sex.

Comment on the patterns you observe in the adjusted variable plot for each scenario.

Use the hints below to run the simulation procedure:

```{r sims, warning = FALSE, message = FALSE}
# Set seed for reproducibility
set.seed(653)
library(ggplot2)
library(dplyr)

data1 <- nepal.anthro |> 
  select(c(1:16)) |> 
  filter(num == 1) |> 
  filter(!is.na(age) & !is.na(arm)) |>
  arrange(age)

# Simulate the scenarios (arm.A, arm.B, arm.C)
data1 <- data1 |>
  mutate(arm.A = 7.8 + 0.03*sex + 0.07*ht + rnorm(nrow(data1), 0, 0.93),
         arm.B = 7.8 + 0.035*age + 0.03*sex + 0.07*ht + rnorm(nrow(data1), 0, 0.93),
         arm.C = 7.8 + 0.05*age - 0.025*(age-6)*(age > 6) + 0.03*sex + 0.07*ht + 
           rnorm(nrow(data1), 0, 0.93))

```


```{r}
#For A
# Obtain residuals R(Y|Z) and R(X|Z)
data1 <- data1 |>
  mutate(resid.armA = lm(arm.C ~ ht + sex, data = data1)$residuals,
         resid.ageA = lm(age ~ ht + sex, data = data1)$residuals)

# Run the model of R(Y|Z) on R(X|Z)
resid.modelA <- lm(resid.armA ~ resid.ageA, data = data1)
summary(resid.modelA)

mrA <- lm(arm ~ age + ht + sex, data = data1)
summary(mrA)

avplotA <- ggplot(data = data1,
                  aes(x = resid.ageA, y = resid.armA)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(aes(y = predict(resid.modelA)), method = "lm", formula = y ~ x,
              linewidth = 1, color = "#024873", se = FALSE) +
  labs(x = "Residuals of age on height, weight, age",
       y = "Residuals of arm circumference on height, age",
       title = "Adjusted variable plot for arm circumference on age \n adjusting for height, age") +
  custom_theme
avplotA
```

```{r}
#for B
# Obtain residuals R(Y|Z) and R(X|Z)
data1 <- data1 |>
  mutate(resid.armB = lm(arm.B ~ ht + sex, data = data1)$residuals,
         resid.ageB = lm(age ~ ht + sex, data = data1)$residuals)

# Run the model of R(Y|Z) on R(X|Z)
resid.modelB <- lm(resid.armB ~ resid.ageB, data = data1)
summary(resid.modelB)

mrB <- lm(arm ~ age + ht + sex, data = data1)
summary(mrB)

avplotB <- ggplot(data = data1,
                  aes(x = resid.ageB, y = resid.armB)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(aes(y = predict(resid.modelB)), method = "lm", formula = y ~ x,
              linewidth = 1, color = "#024873", se = FALSE) +
  labs(x = "Residuals of age on height, weight, age",
       y = "Residuals of arm circumference on height, age",
       title = "Adjusted variable plot for arm circumference on age \n adjusting for height, age") +
  custom_theme
avplotB
```


```{r sims, warning = FALSE, message = FALSE}
# Set seed for reproducibility
set.seed(653)
library(ggplot2)
library(dplyr)

data1 <- nepal.anthro |> 
  select(c(1:16)) |> 
  filter(num == 1) |> 
  filter(!is.na(age) & !is.na(arm)) |>
  arrange(age)

# Simulate the scenarios (arm.A, arm.B, arm.C)
data1 <- data1 |>
  mutate(arm.A = 7.8 + 0.03*sex + 0.07*ht + rnorm(nrow(data1), 0, 0.93),
         arm.B = 7.8 + 0.035*age + 0.03*sex + 0.07*ht + rnorm(nrow(data1), 0, 0.93),
         arm.C = 7.8 + 0.05*age - 0.025*(age-6)*(age > 6) + 0.03*sex + 0.07*ht + 
           rnorm(nrow(data1), 0, 0.93))

```

With the same data above, using the observed age (ð‘‹), sex (ð‘1) height(ð‘2), simulate the arm circumference value (ð‘Œ)

```{r}
#For C
#arm.C = 7.8 + 0.05*age - 0.025*(age-6)*(age > 6) + 0.03*sex + 0.07*ht + 
           #rnorm(nrow(data1), 0, 0.93))

# Obtain residuals R(Y|Z) and R(X|Z)
data1 <- data1 |>
  mutate(resid.armC = lm(arm.C ~ ht + sex, data = data1)$residuals,
         resid.ageC = lm(age ~ ht + sex, data = data1)$residuals)

# Run the model of R(Y|Z) on R(X|Z)
resid.modelC <- lm(resid.armC ~ resid.ageC, data = data1)
summary(resid.modelC)

mrC <- lm(arm ~ age + ht + sex, data = data1)
summary(mrC)

avplotC <- ggplot(data = data1,
                  aes(x = resid.ageC, y = resid.armC)) +
  geom_jitter(alpha = 0.8) +
  geom_smooth(aes(y = predict(resid.modelC)), method = "lm", formula = y ~ x,
              linewidth = 1, color = "#024873", se = FALSE) +
  labs(x = "Residuals of age on height, age",
       y = "Residuals of arm circumference on height, sex",
       title = "Adjusted variable plot for arm circumference on age \n adjusting for height, sex") +
  custom_theme
avplotC
```

